---
title: "Tank bromeliad habitat features serve as predictors of invertebrate abundance and richness"
author: "Theresa (Tess Konnovitch), Gonzalez Lab, RUC - CCIB"
date: "4/29/2021"
output:
    github_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(MuMIn)
require(tidyverse)
require(ggfortify)
require(ggplot2)
require(ResourceSelection)
require(mgcv)
require(car)
```


## Introduction
Invertebrates are responsible for a plethora of important ecosystem services, such as decomposition, water filtration, nutrient cycling, disease mitigation, and more (Lavaelle et al., 2006). Given these invaluable services, the recently observed patterns of invertebrate community and population declines in abundance and richness are causing concern and an increased number of investigations in the scientific community. Classical ecological patterns suggest that invertebrate diversity and abundance are positively correlated with habitat size, although studies have reported system-dependent results (Joque & Field, 2014). Tank bromeliads (Bromeliaceae) provide a unique type of habitat to many invertebrates in the neotropics. Tank bromeliad leaves hold freshwater, creating a microecosystem for invertebrates and an ideal model system for environmental manipulation experiments, due to size and tractability (Srivastava et al., 2004). Using invertebrate response data collected in Brazil, I ran a statistical analysis on three habitat variables (number of leaves, plant area, and maximum volume of water the plant holds) to investigate the relationship between tank bromeliad structure and invertebrate response. Using generalized linear models, I found that these features can predict invertebrate richness and abundance, further supporting the theory that habitat can predict invertebrate community assembly. This study also provides invaluable insight as I prepare for a similar experiment in Brazil.  

## Data
This data was collected by Mark Nessel, along with other members and colleagues of the Gonzalez Lab, in 2019 along the coast of Brazil. Invertebrate communities from 12 plants at each of the 10 sites were analyzed. 

The data frame (pd) contains a row for each plant in the survey. The following are descriptions for the columns referenced in this project:

- `plant_area`: The area of plant, calculated using multiple diameter measurements.
- `Leaves`: Total number of leaves on the plant
- `Max.volume`: Maximum volume in plant
- `id`: refernce ID for the project, created using site name + plant number (e.g., Cardoso1 refers to plant 1 from Cardoso). There are 12 plants from each of the 10 sites. 
- `inv_abundance`: number of individual invertebrates per plant
- `n_species`: number of unique morphospecies per plant

```{r data prep}
setwd("C:/Users/tessa/OneDrive/Documents/Classes/AEDA/konnovitch_theresa")
pd <- read.csv("Survey_Plant_Data.csv")
```

## Methods

My methods consist of five steps: plotting to visualize the data, running statistics, checking assumptions, interpreting the results, and plotting the model back on the data picture. 

### Step 1: Plot the Data
I started by plotting variables to "eyeball" the data. I plotted all combinations of variables (e.g., explanatory vs. response, response vs. response, explanatory vs. explanatory) as a data exploration exercise. 
```{r plots}
#plant area and abundance
ggplot(pd, aes(x = plant_area, y = inv_abundance))+
    geom_point()+
    geom_smooth(span = 1, color = "red", method = "loess")
#max volume and abundance
ggplot(pd, aes(x = Max.volume, y = inv_abundance))+
    geom_point()+
    geom_smooth(span = 1, color = "red", method = "loess")
#leaves and abundance
ggplot(pd, aes(x = Leaves, y = inv_abundance))+
    geom_point()+
    geom_smooth(span = 1, color = "red", method = "loess")
#plant area and richness
ggplot(pd, aes(x = plant_area, y = n_species))+
    geom_point()+
    geom_smooth(span = 1, color = "red", method = "loess")
#max volume and richness
ggplot(pd, aes(x = Max.volume, y = n_species))+
    geom_point()+
    geom_smooth(span = 1, color = "red", method = "loess")
#leaves and richness
ggplot(pd, aes(x = Leaves, y = n_species))+
    geom_point()+
    geom_smooth(span = 1, color = "red", method = "loess")
#leaves and plant area
ggplot(pd, aes(x = Leaves, y = plant_area))+
    geom_point()+
    geom_smooth(span = 1, color = "red", method = "loess")
#leaves and max volume
ggplot(pd, aes(x = Leaves, y = Max.volume))+
    geom_point()+
    geom_smooth(span = 1, color = "red", method = "loess")
#ax volume and plant area
ggplot(pd, aes(x = Max.volume, y = plant_area))+
    geom_point()+
    geom_smooth(span = 1, color = "red", method = "loess")
#richness and abundance
ggplot(pd, aes(x = inv_abundance, y = n_species))+
    geom_point()+
    geom_smooth(span = 1, color = "red", method = "loess")
```



### Step 2: Run Stats

### A) Evaluate correlation

Using the "cor" function, I checked the correlation between variables. This test showed that each explanatory variable is positively correlated with each response variable, although the correlation between plant area and invertebrate abundance, and max volume and species richness are not significantly correlated. Abundance and richness were most correlated with number of leaves. This is consistent the popular technique of using leaf biomass as a predictor of various invertebrate responses, as seen in tank bromeliad literature. Max volume was significantly and positively correlated with both number of leaves and plant area, but leaves and plant area were not. Invertebrate abundance and species richness were significantly and positively correlated, consistent with evidence that suggests richness is influenced by abundance of a sample (Roswell et al., 2021). 

```{r colinear}
cor(pd$inv_abundance, pd$plant_area)
cor.test(pd$inv_abundance, pd$plant_area) #.1694424, not significant

cor(pd$inv_abundance, pd$Max.volume)
cor.test(pd$inv_abundance, pd$Max.volume) #.2293003, significant

cor(pd$inv_abundance, pd$Leaves)
cor.test(pd$inv_abundance, pd$Leaves) #.3415183, significant

cor(pd$n_species, pd$plant_area)
cor.test(pd$n_species, pd$plant_area) #0.09465695, no significant

cor(pd$n_species, pd$Max.volume)
cor.test(pd$n_species, pd$Max.volume) #0.128345, not significant

cor(pd$n_species, pd$Leaves)
cor.test(pd$n_species, pd$Leaves) #0.2643559, significant

cor(pd$Max.volume, pd$Leaves)
cor.test(pd$Max.volume, pd$Leaves) #0.3748002, significant

cor(pd$Max.volume, pd$plant_area)
cor.test(pd$Max.volume, pd$plant_area) #0.3191787, significant

cor(pd$Leaves, pd$plant_area)
cor.test(pd$Leaves, pd$plant_area) #.1601275, not significant

cor(pd$n_species, pd$inv_abundance)
cor.test(pd$n_species, pd$inv_abundance) #0.3789065, significant

```


### B) Evaluate Predictors

I used generalized linear modeling to determine if the explanatory variables predicted invertebrate responses, due to the dataset being "count" data that seemed relatively linear. I initially thought to use generalized additive models, given the slight curvature of the data picture. The GAM output showed an edf of 1, indicating that the relationship is linear. I then compared both GAM and GLM models to find the best model. The number of leaves + plant area was the best predictor for invertebrate abundance, and number of leaves + volume was the best model for invertebrate richness. 

```{r correlogram}
#models for abundance
abundance <- log(pd$inv_abundance)

ab_A <- glm(abundance ~ plant_area, data = pd)
ab_V <- glm(abundance ~ Max.volume, data = pd)
ab_L <- glm(abundance ~ Leaves, data = pd)
ab_VA <- glm(abundance ~ plant_area+Max.volume, data = pd)
ab_LV <- glm(abundance ~ Max.volume+Leaves, data = pd)
ab_AL <- glm(abundance ~ Leaves+plant_area, data = pd)
ab_VAL <- glm(abundance ~ Leaves+plant_area+Max.volume, data = pd)
ab_A2 <- gam(abundance ~ plant_area, data = pd)
ab_V2 <- gam(abundance ~ Max.volume, data = pd)
ab_L2 <- gam(abundance ~ Leaves, data = pd)
ab_VA2 <- gam(abundance ~ plant_area+Max.volume, data = pd)
ab_LV2 <- gam(abundance ~ Max.volume+Leaves, data = pd)
ab_AL2 <- gam(abundance ~ Leaves+plant_area, data = pd)
ab_VAL2 <- gam(abundance ~ Leaves+plant_area+Max.volume, data = pd)

model.sel(ab_A, ab_V, ab_L, ab_VA, ab_LV, ab_AL, ab_VAL, ab_A2, ab_V2, ab_L2, ab_VA2, ab_LV2, ab_AL2, ab_VAL2) #ab_AL = best model


#models for richness
r_A <- glm(n_species ~ plant_area, data = pd)
r_V <- glm(n_species ~ Max.volume, data = pd)
r_L <- glm(n_species ~ Leaves, data = pd)
r_VA <- glm(n_species ~ plant_area+Max.volume, data = pd)
r_LV <- glm(n_species ~ Max.volume+Leaves, data = pd)
r_AL <- glm(n_species ~ Leaves+plant_area, data = pd)
r_VAL <- glm(n_species ~ Leaves+plant_area+Max.volume, data = pd)
r_A2 <- glm(n_species ~ plant_area, data = pd)
r_V2 <- glm(n_species ~ Max.volume, data = pd)
r_L2 <- glm(n_species ~ Leaves, data = pd)
r_VA2 <- glm(n_species ~ plant_area+Max.volume, data = pd)
r_LV2 <- glm(n_species ~ Max.volume+Leaves, data = pd)
r_AL2 <- glm(n_species ~ Leaves+plant_area, data = pd)
r_VAL2 <- glm(n_species ~ Leaves+plant_area+Max.volume, data = pd)

model.sel(r_A, r_V, r_L, r_VA, r_LV, r_AL, r_VAL,r_A2, r_V2, r_L2, r_VA2, r_LV2, r_AL2, r_VAL2) 
#r_LV = best

```


### Step 3: Check Assumptions
Invertebrate abundance was log transformed before creating the models, so that the AIC values were from the same dataset, and therefore comparable. Both models show normality based on the histogram of residuals, autoplot, and summary. The histogram of residuals shows symmetric distribution around 0. The QQplot also suggests that residuals are normal.The variance inflation factor function (vif) was used to determine the extent of correlation between predictors in the best models. The VIF output was below five, indicating that multicollinearity is not a problem in either model. 

```{r standardize}
#abundance
hist(ab_AL$residuals)
autoplot(ab_AL)
summary(ab_AL) 
vif(ab_AL)


#richness
hist(r_L$residuals)
autoplot(r_LV)
summary(r_LV)
vif(ab_AL)

```





## Results

### Step 4: Interpret Results

The summary of both models show that the residuals have symmetric distribution across zero, indicating normality and that the model fits the data well. The standard errors are less than the respective coefficient, indicating low variation. For abundance, the residual deviance had reduced 12.71 with a loss of 1 degree of freedom. For richness, the residual deviance has reduced 41.82 with a loss of 1 degree of freedom. Since deviance indicates the "goodness of fit," this shows that the models fit the data better than the null model. I also decided to confirm this by using the Hosmer-Lemeshow Goodness of Fit test (Lillis, 2021), which found that there was no significant difference between the model and observed data (p value > 0.05). The Fisher Scoring iterations show that the model did converge. 

```{r glmfit}
summary(r_LV)
summary(ab_AL)

hoslem.test(pd$inv_abundance, fitted(ab_AL))
hoslem.test(pd$n_species, fitted(ab_LV))
```


## Plot Model

The last step of my methodology was to plot the newly formed models back on the data picture. Each figure shows the original data (black dots), the model (red line) and 95% confidence intervals (transparent red ribbon). 

```{r glmres}
#abundance (Leaves)

a_intercept = coef(ab_AL)[1]
a_slope = coef(ab_AL)[2]
and <- data.frame(Leaves = seq(1,50, by=1), plant_area = seq(1, 50, by = 1)) # make a data.frame of new explanatory variables
and$ab_AL <- predict(ab_AL, newdata=and, type='response') # predict on the scale of the response
and$ase <- predict(ab_AL, newdata=and, type='response', se.fit=TRUE)$se.fit # get the standard errors of the fit

ggplot(pd, aes(x=Leaves, y=abundance)) +
    geom_point() +
    geom_line(data=and, aes(x=Leaves, y=ab_AL, color='red')) +
    geom_ribbon(data=and, aes(x=Leaves, ymin=ab_AL-ase, ymax=ab_AL+ase),alpha=0.3, fill='red', inherit.aes=FALSE) +
    ggtitle("Best Generalized Linear Model for Invertebrate Abundance") +
    xlab("Number of Leaves per Plant") +
    ylab("Number of Individuals per Plant") +
    scale_colour_discrete("Leaves + Plant Area Model")

#abundance (plant area)

a_intercept = coef(ab_AL)[1]
a_slope = coef(ab_AL)[2]
and <- data.frame(Leaves = seq(1,50, by=1), plant_area = seq(1, 50, by = 1)) # make a data.frame of new explanatory variables
and$ab_AL <- predict(ab_AL, newdata=and, type='response') # predict on the scale of the response
and$ase <- predict(ab_AL, newdata=and, type='response', se.fit=TRUE)$se.fit # get the standard errors of the fit

ggplot(pd, aes(x=Leaves, y=abundance)) +
    geom_point() +
    geom_line(data=and, aes(x=plant_area, y=ab_AL, color='red')) +
    geom_ribbon(data=and, aes(x=plant_area, ymin=ab_AL-ase, ymax=ab_AL+ase),alpha=0.3, fill='red', inherit.aes=FALSE) +
    ggtitle("Best Generalized Linear Model for Invertebrate Abundance") +
    xlab("Plant Area") +
    ylab("Number of Individuals per Plant")  +
    scale_colour_discrete("Leaves + Plant Area Model")

#richness (Leaves)

r_intercept = coef(r_LV)[1]
r_slope = coef(r_LV)[2]
rnd <- data.frame(Leaves = seq(1,50, by=1), Max.volume = seq(0.1,5, by = 0.1)) # make a data.frame of new explanatory variables
rnd$r_LV <- predict(r_LV, newdata=rnd, type='response') # predict on the scale of the response
rnd$rse <- predict(r_LV, newdata=rnd, type='response', se.fit=TRUE)$se.fit # get the standard errors of the fit

ggplot(pd, aes(x=Leaves, y=n_species)) +
    geom_point() +
    geom_line(data=rnd, aes(x=Leaves, y=r_LV, color='red')) +
    geom_ribbon(data=rnd, aes(x=Leaves, ymin=r_LV-rse, ymax=r_LV+rse),alpha=0.3, fill='red', inherit.aes=FALSE) +
    ggtitle("Best Generalized Linear Model for Invertebrate Species Richness") +
    xlab("Number of Leaves per Plant") +
    ylab("Number of Species per Plant")  +
    scale_colour_discrete("Leaves + Max Volume Model")

#Richness (Max.volume)

r_intercept = coef(r_LV)[1]
r_slope = coef(r_LV)[2]
rnd <- data.frame(Leaves = seq(1,50, by=1), Max.volume = seq(0.1,5, by = 0.1)) # make a data.frame of new explanatory variables
rnd$r_LV <- predict(r_LV, newdata=rnd, type='response') # predict on the scale of the response
rnd$rse <- predict(r_LV, newdata=rnd, type='response', se.fit=TRUE)$se.fit # get the standard errors of the fit

ggplot(pd, aes(x=Max.volume, y=n_species)) +
    geom_point() +
    geom_line(data=rnd, aes(x=Max.volume, y=r_LV, color='red')) +
    geom_ribbon(data=rnd, aes(x=Max.volume, ymin=r_LV-rse, ymax=r_LV+rse),alpha=0.3, fill='red', inherit.aes=FALSE) +
    ggtitle("Best Generalized Linear Model for Invertebrate Species Richness") +
    xlab("Max Volume per Plant") +
    ylab("Number of Species per Plant") +
    scale_colour_discrete("Leaves + Max Volume Model")

```


## Discussion

As discussed throughout the methods, this analysis shows that certain habitat features (number of leaves, plant area, and max volume) are correlated with and can serve as predictors for both invertebrate community abundance and species richness in Brazilian tank bromeliads. These models were indicated to fit the data well, using the series of steps and tests above. This relationship should be taken into account when using tank bromeliads as model systems to investigate invertebrate responses to changes in environment, among other studies. 


## Future Research

While only in my first few months of my Ph.D., I have yet to collect any field data. However, I plan to venture to Brazil to conduct a similar experiment next winter. This statistical analysis has increased my awareness of the impact of certain habitat features on invertebrate responses. These results will certainly be implemented in my future research, which investigates the effects of nutrient availability and increasing temperature on invertebrate community structure. 


## Work Cited

Dézerald, O., Leroy, C., Corbara, B., Dejean, A., Talaga, S., & Céréghino, R. (2017). Environmental drivers of invertebrate population dynamics in Neotropical tank bromeliads. Freshwater Biology, 62(2), 229–242. https://doi.org/10.1111/fwb.12862

Dézerald, O., Leroy, C., Corbara, B., Dejean, A., Talaga, S., & Céréghino, R. (2018). Tank bromeliads sustain high secondary production in neotropical forests. Aquatic Sciences, 80(2), 1–12. https://doi.org/10.1007/s00027-018-0566-3

Jocque, M., & Field, R. (2014). Aquatic invertebrate communities in tank bromeliads: How well do classic ecological patterns apply? Hydrobiologia, 730(1), 153–166. https://doi.org/10.1007/s10750-014-1831-7

Jocque, M., Kernahan, A., Nobes, A., Willians, C., & Field, R. (2010). How effective are non-destructive sampling methods to assess aquatic invertebrate diversity in bromeliads? Hydrobiologia, 649(1), 293–300. https://doi.org/10.1007/s10750-010-0272-1

Lavelle, P., Decaëns, T., Aubert, M., Barot, S., Blouin, M., Bureau, F., Margerie, P., Mora, P., & Rossi, J. P. (2006). Soil invertebrates and ecosystem services. European Journal of Soil Biology, 42(SUPPL. 1). https://doi.org/10.1016/j.ejsobi.2006.10.002

Rosewell, M., Dushoff, J., & Winfree, R. (2021). A conceptual guide to measuring species diversity. OIKOS, 130:321-338. doi: 10.1111/oik.07202

Srivastava, D. S., Kolasa, J., Bengtsson, J., Gonzalez, A., Lawler, S. P., Miller, T. E., Munguia, P., Romanuk, T., Schneider, D. C., & Trzcinski, M. K. (2004). Are natural microcosms useful model systems for ecology? Trends in Ecology and Evolution, 19(7), 379–384. https://doi.org/10.1016/j.tree.2004.04.010

